A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-03 paint-contour-tree.tex
minor changes

\newcommand{\atgen}{\symbol{'100}}
\newcommand{\si}[1]{#1}
\newcommand{\BenThanks}[1]{\thanks{%
      Department of Computer Science; %
      University of Illinois; %
      201 N. Goodwin Avenue; %
      Urbana, IL, 61801, USA; %
      {\tt \si{raichel}2\atgen{}\si{uiuc}.\si{edu}}; %
      {\tt \url{\si{http://illinois.edu/\string~\si{raichel2}}}}%
      . %
      #1}}

\author{
  Benjamin Raichel\BenThanks{}
  \and
  C. Seshadhri
}

@@

\Fig{relative} and \Fig{sorting} show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. They correspond to contours (which are $(d-1)$-manifolds) gaining or losing handles.

@@

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching \Thm{main-alg}.

\begin{definition}\label{def:hfdt}
A \emph{height function decision tree} is an algorithm
that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
tree over the heights. The running time is simply the number of queries made.
\end{definition}

\begin{theorem} \label{thm:main-lb} 
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider the set of functions $\bF_C$ that all same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

@@

Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
of inputs of fixed cost.

\paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
computing contour trees is expensive, and provide insight into the hardness of particular instances. It is intriguing
that the cost (as defined earlier) roughly captures the hardness of an instance.
In that sense, it is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

@@

\Ben{This is already defined in section 3, should we use the same numbering?}
\begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$.

@@

\section{Lower Bound by Path Decomposition}
\label{sec:lb}
In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
we first consider the easier case of join trees, after which the lower bound for more general contour trees will easily follow.
\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.

>>>>

\newcommand{\atgen}{\symbol{'100}}
\newcommand{\si}[1]{#1}
\newcommand{\BenThanks}[1]{\thanks{%
      Department of Computer Science; %
      University of Illinois; %
%      201 N. Goodwin Avenue; %
      Urbana, IL, USA; %
      {\tt \si{raichel}2\atgen{}\si{uiuc}.\si{edu}}; %
      {\tt \url{\si{http://illinois.edu/\string~\si{raichel2}}}}%
      . %
      #1}}

\author{
  Benjamin Raichel\BenThanks{}
  \and
  C. Seshadhri\thanks{Sandia National Laboratories; Livermore, CA, USA; {\tt scomand@sandia.gov}}
}

@@

\Fig{relative} and \Fig{sorting} show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
For example, these could be events where contours gain or lose handles.

@@

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching the performance of \Thm{main-alg}. 
Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
of inputs of fixed cost.

%\begin{definition}\label{def:hfdt}
%A \emph{height function decision tree} is an algorithm
%that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
%tree over the heights. The running time is simply the number of queries made.
%\end{definition}
%
\begin{theorem} \label{thm:main-lb} 
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

@@

\paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
computing contour trees is expensive, and provide insight into the hardness of particular instances. It is intriguing
that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fast are easy to compute, whereas trees with long paths are hard.
Our result is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

\section{Acknowledgements}

We thank Jeff Erickson and Yusu Wang for numerous useful discussions. This work is
supported by the Laboratory Directed Research and Development
(LDRD) program of Sandia National Laboratories. Sandia National
Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed
Martin Corporation, for the U.S. Department of Energy’s National
Nuclear Security Administration under contract DE-AC04-94AL85000.

@@

%\Ben{This is already defined in section 3, should we use the same numbering?}
\begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$.

@@

\section{Lower Bound by Path Decomposition}
\label{sec:lb}
In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  
We first prove a lower bound for join trees, and then generalize to contour trees.
%\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.
